Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects

### 2025-05-01 paper.qmd
Reviewed until C. Impact probability density for debris of inclination i

With forecasts predicting a steadily growing number of flights, see @fig-forecast, and the modern
race to commercialise space call for a re-evaluation of the risk of
collision and casualties due to uncontrolled debris re-entries.
Overseeing authorities need methodologies and tools to assess which 
safety measures are the most adequate to minimize casualties and economic
impact.

@@

This paper develops a practical and straightforward pipeline to rapidly
assess the risk from uncontrolled re-entry of space debris for the European
region.
Apart from the general risk assessment of potential re-entries, we analyse
a soon to occur re-entry of debris from a 1972 failed USSR mission to Venus.

@@

# ~constellation, ~state, ~launched, ~active, ~planned, ~first_launch,
megas |> 
  gt() |> 
   cols_label(
    constellation ~ "Constellation",
    launched ~ "Launched",
    active ~ "Active",
    planned ~ "Planned",
    first_launch ~ "First launch"
  ) |> 
   sub_missing(
    columns = "first_launch",
    missing_text = "-"
  ) |> 
  opt_table_font(size = "10px")

@@

We follow the same mathematical approach of 
[@patera2008; @wrightAirspaceClosuresDue2025] but instead of
approximating flight density with population density [@patera2008] or
aircraft transponder data [@wrightAirspaceClosuresDue2025], we use flown 
trajectories.

We base our hourly flight density estimation on flown \acr{IFR} trajectories as
recorded by EUROCONTROL's \acr{NM}[^1] and binned on Uber's H3 gridding system
[@uber-h3] for fast processing.  
This work will use H3 cells at resolution 3 which corresponds to hexagons of
average area $\qty{12393}{km^2}$ and average edge length of
$\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.

@@

H3 cells (red) at resolution 3 and their centers (blue) covering the bounding
box of 43 EUROCONTROL Member States FIR's (blue shaded polygon).

@@

We characterise densities on an hourly basis for the days of the 2023 and 2024
busiest \acr{IATA} summer ISO week.

We expand on the works of [@wrightAirspaceClosuresDue2025; @patera2008] and
provide a reproducible example for States, Agencies or relevant authorities
to build upon.

@@

Next sections will develop the mathematcal background for the evaluation of
collision and casuality risk of which traffic density is a key component.

@fig-density-summer2024-res-3 shows hourly flight density for 2024-07-05.
Expected outcomes such as cells at zero density for war or no-fly zones
(Ukraine, Libia or Syria) or typical traffic flows (North Atlantic, Canaries
and East-bound over Türkiye) are clearly highlighted.

@@

Flight density in the EUROCONTROL Area on 2024-07-05, the busiest day of the busiest week in 2024 (H3 cells at resolution $3$).
:::

## Impact Probability for Single Flight, Single Debris
The probability, $p_k$, of a debris (point) object $k$, impacting an
aircraft (point) object over the area, $a$, is expressed as:

@@

If we discretaze the area of interest $a$ using H3 cells,
we have

$$
p_k = \varepsilon_k \sum_h \rho(h) \sigma_k(h)
$$

So for cell $h$ we have

$$
p_{k}(h) = \varepsilon_k \rho(h) \sigma_k(h)
$$

@@

We assume the aircraft impact area to be much larger than the debris
object area so as to neglet the debris impact area.
We can also simplify the geometry of the aircraft and model it as
a rectangular solid with dimesions of wingspan, length and height. 
Now, if the aircraft were stationary, the area exposed to a debris impact would
be the top view, while if the debris were stationary it would be the front view.
In reality the aircraft is moving (manly) horizontally and the debris 
vertically [^balistic_descent], which which allows to express the effective
exposed area, $\varepsilon_k$, as

@@

In our calculations we will use ($t$ subscript for aircraft type)

@@

[^balistic_descent]: *Even for a ballistic coefficient of 50 lbs/ft2, the flight
path angle of a reentering object at 40,000 feet is -86.6 degrees, which
corresponds to nearly vertical descent.
Since most debris objects have ballistic coefficients equal to or 
less than 50 lbs/ft2, the vertical descent assumption is a very
accurate approximation.* [@patera2008]

>>>>

With forecasts predicting a steadily growing number of flights, see @fig-forecast, and the modern
race to commercialise space it is clear a re-evaluation of the risks of
collision and casualties due to uncontrolled debris re-entries is needed.
Overseeing authorities need methodologies and tools to assess which 
safety measures are the most adequate to minimise potential casualties and economic
impact.

@@

This paper develops a practical and straightforward pipeline to rapidly
assess the risk from uncontrolled re-entry of space debris for the European
region.
Apart from the general risk assessment of potential re-entries, 
a soon to occur re-entry of debris from a 1972 failed USSR mission to Venus is analysed.

@@

# ~constellation, ~state, ~launched, ~active, ~planned, ~first_launch,
megas |> 
  gt() |> 
   cols_label(
    constellation ~ "Constellation",
    state ~ "Country",
    launched ~ "Launched",
    active ~ "Active",
    planned ~ "Planned",
    first_launch ~ "First launch"
  ) |> 
   sub_missing(
    columns = "first_launch",
    missing_text = "-"
  ) |> 
  opt_table_font(size = "10px")

@@

We follow the same mathematical approach of 
[@patera2008; @wrightAirspaceClosuresDue2025] but instead of
approximating flight density with population density [@patera2008] or
aircraft transponder data [@wrightAirspaceClosuresDue2025], we use actual flown
flight trajectories.

The hourly flight density estimation is based on on flown \acr{IFR} trajectories as
recorded by EUROCONTROL's \acr{NM}[^1] and binned using Uber's H3 geospatial indexing system
[@uber-h3] in order to decrease spatial complexity which in turn enables fast processing.
This work will use H3 cells at resolution 3 which corresponds to hexagons of an
average area $\qty{12393}{km^2}$ and an average edge length of
$\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.

@@

H3 cells (red) at resolution 3 and their centers (blue) covering the bounding
box of 43 EUROCONTROL Member States \acr{FIR}s (blue shaded polygon).

@@

The flight densities are characterised on an hourly basis for all days in the busiest \acr{IATA} summer ISO week in 2023 and 2024.

This paper expands the works of [@wrightAirspaceClosuresDue2025; @patera2008] and
provides a reproducible example for States, Agencies or relevant authorities
to build upon.

@@

Next sections will develop the mathematical background for the evaluation of
collision and casualties risk of which traffic density is a key component.

@fig-density-summer2024-res-3 shows hourly flight density for the 5th of July 2024.
Expected outcomes such as cells at zero density for war or no-fly zones
(Ukraine, Libia or Syria) or typical traffic flows (North Atlantic, Canaries
and East-bound over Türkiye) are clearly highlighted.

@@

Flight density in the EUROCONTROL Area on the 5th of July 2024, the busiest day of the busiest week in 2024 (H3 cells at resolution $3$).
:::

## Impact probability for a single flight and single debris
The probability, $p_k$, of a debris (point) object $k$, impacting an
aircraft (point) object over the area, $a$, is expressed as:

@@

Using H3 to discretise the area of interest $a$ using H3 cells gives

$$
p_k = \varepsilon_k \sum_h \rho(h) \sigma_k(h).
$$

Where for each cell $h$ the following applies

$$
p_{k}(h) = \varepsilon_k \rho(h) \sigma_k(h).
$$

@@

It is assumed the aircraft impact area to be much larger than the debris
object area so as to neglect the debris impact area.
The geometry of the aircraft is simplified and modelled as
a rectangular cuboid with dimensions of wingspan, length and height. 
Now, if the aircraft were stationary, the area exposed to a debris impact would
be the top view, while if the debris were stationary it would be the front view.
In reality an aircraft en-route is moving (mainly) horizontally and the debris 
vertically [^balistic_descent], which which allows to express the effective
exposed area, $\varepsilon_k$, as

@@

In the calculations the following is assumed to be true ($t$ subscript for aircraft type)

@@

[^balistic_descent]: *Even for a ballistic coefficient of 50 lbs/ft2, the flight
path angle of a re-entering object at 40,000 feet is -86.6 degrees, which
corresponds to nearly vertical descent.
Since most debris objects have ballistic coefficients equal to or 
less than 50 lbs/ft2, the vertical descent assumption is a very
accurate approximation.* [@patera2008]
